Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-26 manuscript/manuscript.tex
Manuscript updated

\emph{Manuscript last updated: 2025-07-08 18:15:17.700176}

@@

However, the comparability and transportability of results derived
between non-interventional studies and RCTs depends on multiple factors.
Frequently referenced limitations include lack of baseline randomization
and the imbalance in prognostic factors resulting thereof, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{3,4} adoption of changes in treatment
guidelines over time in routine clinical care and the inability to
measure and emulate common eligibility criteria, including prognostic
factors, and standardized response assessments in real-world data
(RWD).\textsuperscript{5} While examples of oncology trial emulations
have been published,\textsuperscript{5--7} a systematic and scaled
approach to emulate a diverse set of oncology trials with multiple
heterogeneous databases is necessary to increase confidence in the
interpretability of non-interventional studies, evaluate regulatory
considerations, and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{8} increased our
understanding of when non-interventional studies can come to causal
conclusions on treatment effects by benchmarking results against RCTs
under the assumption that each RCT finding reflects a causal treatment
effect. In settings where the RCT designs could be emulated well,
non-interventional studies were able to reach similar
conclusions.\textsuperscript{9} However, this prior work from
RCT-DUPLICATE focused primarily on emulating trials in non-oncology
settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts
non-interventional studies can or cannot yield similar results compared
to oncology RCTs. Therefore, the project emphasizes trials of therapies
for common cancers for which there has been substantial therapeutic
development in recent years. After review and collaboration with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple accessible databases (i.e., the total
equaling 12 trial emulations x \emph{n} databases which are found
fit-for-purpose for each trial).

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: Parallel assignment
\item
  Industry-sponsored
\item
  Trial start: 2011 or later
\item
  Primary purpose: Study treatment efficacy
\item
  Endpoint(s): Overall survival must be one of the endpoints reported
  (either as hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization for each criterion is listed in
detail in Table~\ref{tbl-criteria}. We mainly considered pivotal large
late-phase RCTs that were initiated after 2011 because treatment
guidelines for included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of newly-approved
therapies in routine care, patients are less likely to receive outdated
treatment regimens in clinical practice. Conversely, we excluded trials
with results published too recently to allow for substantive medical
product uptake and follow-up time accrual in real-world databases used
for this project. We did not define a global cut-off as the requirements
for follow-up time are different for each cancer type and population
(e.g., advanced NSCLC versus early-stage breast cancer) and the
decisions were made on a clinically relevant basis.

@@

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
up candidate trials are listed in \textbf{Supplementary Table 1}. The
majority of shortlisted trials represent indications for advanced
(locally advanced, inoperable, recurrent/progressive disease) or
metastatic cancer patients because a large proportion of drug
development efforts have recently focused on these settings. A key
objective that we aim to explore with the selected trials is to evaluate
how stage (early, late), line of therapy ({[}neo{]}adjuvant, first line,
advanced lines of therapy), therapy protocols (monotherapy, combination
therapy) and clinical characteristics (all-comer versus complex genetic
or immunological signatures) can be emulated using RWD. If more thorough
feasibility assessments after applying all relevant eligibility criteria
suggest that the threat of bias from mis-measurement of key study
parameters or residual confounding remains high, or that the study size
is not sufficient, then runner-up candidate trials will be considered
instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific
EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA,
Flatiron Health, and Ontada/McKesson. A detailed description and
sampling methodology will be provided with each trial emulation
protocol. For ENCORE, not all databases will be available for each
cancer indication given the specificity of data elements required and
the names of the databases will be blinded and referred to as ENCORE
DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in
randomly assigned order). If more than one database is considered
fit-for-use for a respective trial emulation, the most suitable analytic
model will be employed for each database separately and final treatment
effect estimates will be pooled using a meta-analytic approach using
fixed effects (weights reflect the inverse of the variance of each
database study) and random effects (weights reflect the inverse of the
variance of each database study plus an additional variance term that
quantifies the assumed heterogeneity between databases)
models.\textsuperscript{20}

@@

The distribution of patient characteristics, by exposure status, will be
examined before and after applying eligibility criteria and contrasted
with the distributions of patient characteristics of the original RCT.
Initial propensity score matching or weighting methods will be applied
to ensure that measured pre-exposure covariates can be balanced,
exposure cohorts are conditionally exchangeable at baseline, and
resulting sample sizes are still sufficient after matching or weighting.
At this stage, all exploratory analyses will be conducted blinded
towards the outcome to not bias any study design or analytic choices
based on known outcome information.

@@

Causal analyses of non-interventional data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data .\textsuperscript{28} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreement metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time from index treatment initiation to all-cause mortality (rwOS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

@@

To formally compare treatment effects between RCTs and their respective
emulations, we will adapt the approach of the RCT-DUPLICATE
project.\textsuperscript{8,30} That is, for the primary endpoint of
interest (hazard ratio {[}HR{]} for OS and corresponding 95\% confidence
intervals), we will derive three qualitative agreement metrics:
statistical agreement, estimate agreement and agreement based on the
standardized mean difference (SMD). Examples are illustrated in
Table~\ref{tbl-metrics}.

@@

where \(\hat{\theta}_{RCT}\) and \(\hat{\theta}_{RWE}\) are the
treatment effect estimates (hazard ratios or median survival time
differences) and \(\text{Var}(\hat{\theta}_{RCT})\) and
\(\text{Var}(\hat{\theta}_{RWE})\) are the corresponding variances for
RCT and RWE, respectively. The resulting SMDs will be interpreted such
that with an SMD of 1.00, the effect estimate from the RCT and the RWE
emulation are 1 standard deviation apart. For an \(\alpha\)-level of
0.05, the null hypothesis of no difference would be rejected whenever
\(|Z| > 1.96\).

@@

The study design for each trial emulation will be visualized as part of
the protocol using a graphical depiction of the exact measurement
windows of eligibility criteria, washout periods, and covariates
relative to the cohort entry time.\textsuperscript{31}

@@

To establish an analytic cohort, missingness will be assessed across
patients who meet all eligibility criteria, including those with missing
values as a first iteration. These missing data investigations will
empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
principled process on missing data that empirically evaluates different
aspects across partially observed covariates based on three group
diagnostics.\textsuperscript{33,34} These diagnostics cover (1)
comparisons of patients characteristics with and without an observed
level of the partially observed covariate, (2) ability to predict
missingness given observed data, and (3) assessments if outcomes between
patients with a missing value are systematically different. Expert
domain knowledge and assumptions about the underlying missing data
structure through canonical causal diagrams\textsuperscript{35} will
additionally inform decisions regarding the inclusion or exclusion of
individuals with missing values in key eligibility criteria and
potential sensitivity analyses to assess the robustness of these
decisions.\textsuperscript{36}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it has been shown that especially in the context
of partially observed covariate data (as opposed to missing exposure and
outcome data), generally only mechanisms in which a covariate causes its
own missingness lead to significant bias (MNAR).\textsuperscript{35}
Hence, methodologies which retain otherwise eligible patients and give
the potential to adjust for a broader set of prognostic factors (e.g.,
multiple imputation\textsuperscript{37} or doubly robust
methods\textsuperscript{38}) may be preferred over complete case
analyses, although such tradeoffs must be carefully evaluated within the
clinical and study contexts.

@@

Due to its frequency of use in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal HR coefficient for
the treatment comparison for OS.\textsuperscript{39} We will also
consider alternative endpoints on an absolute risk scale such as median
survival times, survival probabilities at pre-defined time points during
follow-up\textsuperscript{40} and restricted mean survival times as
secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of relevant prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{41} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{42,43} As appropriate, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{44,45} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{42}

To assess the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is often considered
reasonable at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
will compute the average post-matching or post-weighting
C-statistics.\textsuperscript{47} In addition, we will use a published
prognostic score for OS\textsuperscript{41} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{48}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given follow-up
time during.\textsuperscript{49,50} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

@@

Building on an established approach developed by the RCT DUPLICATE
project\textsuperscript{8,30}, the ENCORE project aims to emulate 12
oncology trials using four oncology EHR data sources to inform the
potential use of RWD for regulatory purposes in oncology. The project
focuses on four common cancers and will evaluate the agreement of
treatment effect estimates between RCTs and their respective emulations.
Historically, administrative claims databases have been the predominant
data source used to evaluate medical product safety and effectiveness in
the postmarketing setting. With increasing access to EHR data and
availability of granular clinical data elements often necessary to
gather fit-for-use data in oncology, as well as a maturing set of
methodological approaches for causal inference using such data, the
ENCORE project is one of the first of its kind to evaluate contexts
where RWD may be reliably used to draw similar conclusions compared to
RCTs in the field of oncology.

The RCT DUPLICATE initiative\textsuperscript{8,30} identified multiple
emulation challenges that may be similar in ENCORE. One particular
aspect that we may not always be able to emulate is the exact
distribution of patient characteristics of the trial population. This
can be due to the lack of data granularity and comprehensiveness to
emulate relevant eligibility criteria or the fact that large pivotal
trials in oncology are typically conducted in multiple countries
worldwide (all considered databases reflect the US only). This may
impact the results of this trial emulation given that the distribution
of genetic alterations and biomarkers, prognosis, and factors that drive
heterogeneous treatment outcomes of certain cancers can differ between
countries. For example, certain cancers (e.g., GI cancers) or certain
genetic mutations (e.g., EGFR mutations in lung cancer) are much more
prevalent in certain geographic regions and populations compared to the
US. Additionally, treatment landscapes such as the preference for
specific regimens or compounds, their use in earlier or later lines of
therapies, and supportive care practices can differ between geographic
regions, too.

@@

\bibitem[\citeproctext]{ref-hernan2022causal}
\CSLLeftMargin{28. }%
\CSLRightInline{Hernán, M. A. Causal analyses of existing databases: No
power calculations required. \emph{Journal of clinical epidemiology}
\textbf{144}, 203--205 (2022).}

\bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
\CSLLeftMargin{29. }%
\CSLRightInline{Schoenfeld, D. The asymptotic properties of
nonparametric tests for comparing survival distributions.
\emph{Biometrika} \textbf{68}, 316--319 (1981).}

\bibitem[\citeproctext]{ref-franklin2020nonrandomized}
\CSLLeftMargin{30. }%
\CSLRightInline{Franklin, J. M. \emph{et al.} Nonrandomized real-world
evidence to support regulatory decision making: Process for a randomized
trial replication project. \emph{Clinical Pharmacology \& Therapeutics}
\textbf{107}, 817--826 (2020).}

\bibitem[\citeproctext]{ref-schneeweiss2019graphical}
\CSLLeftMargin{31. }%
\CSLRightInline{Schneeweiss, S. \emph{et al.} Graphical depiction of
longitudinal study designs in health care databases. \emph{Annals of
internal medicine} \textbf{170}, 398--406 (2019).}

\bibitem[\citeproctext]{ref-rubin1976inference}
\CSLLeftMargin{32. }%
\CSLRightInline{Rubin, D. B. Inference and missing data.
\emph{Biometrika} \textbf{63}, 581--592 (1976).}

\bibitem[\citeproctext]{ref-weberpals2024smdi}
\CSLLeftMargin{33. }%
\CSLRightInline{Weberpals, J. \emph{et al.} Smdi: An r package to
perform structural missing data investigations on partially observed
confounders in real-world evidence studies. \emph{JAMIA open}
\textbf{7}, ooae008 (2024).}

\bibitem[\citeproctext]{ref-weberpals2024}
\CSLLeftMargin{34. }%
\CSLRightInline{Weberpals, J. \emph{et al.}
\href{https://doi.org/10.2147/CLEP.S436131}{A principled approach to
characterize and analyze partially observed confounder data from
electronic health records}. \emph{Clinical Epidemiology} \textbf{16},
329--343 (2024).}

\bibitem[\citeproctext]{ref-moreno2018canonical}
\CSLLeftMargin{35. }%
\CSLRightInline{Moreno-Betancur, M. \emph{et al.} Canonical causal
diagrams to guide the treatment of missing data in epidemiologic
studies. \emph{American journal of epidemiology} \textbf{187},
2705--2715 (2018).}

\bibitem[\citeproctext]{ref-tompsett2018use}
\CSLLeftMargin{36. }%
\CSLRightInline{Tompsett, D. M., Leacy, F., Moreno-Betancur, M., Heron,
J. \& White, I. R. On the use of the not-at-random fully conditional
specification (NARFCS) procedure in practice. \emph{Statistics in
medicine} \textbf{37}, 2338--2353 (2018).}

\bibitem[\citeproctext]{ref-weberpals2024hdmi}
\CSLLeftMargin{37. }%
\CSLRightInline{Weberpals, J. \emph{et al.} High-dimensional multiple
imputation (HDMI) for partially observed confounders including natural
language processing-derived auxiliary covariates. \emph{American Journal
of Epidemiology} (2025).at
\textless{}\url{https://arxiv.org/abs/2405.10925}\textgreater{}}

\bibitem[\citeproctext]{ref-Shaw2024}
\CSLLeftMargin{38. }%
\CSLRightInline{Shaw, C. K., P. \& Williamson, B. D. Assessing treatment
effects in observational data with missing confounders: A comparative
study of practical doubly-robust and traditional missing data methods.
\emph{GitHub repository} (2024).at
\textless{}\url{https://github.com/PamelaShaw/Missing-Confounders-Methods}\textgreater{}}

\bibitem[\citeproctext]{ref-cox1972regression}
\CSLLeftMargin{39. }%
\CSLRightInline{Cox, D. R. Regression models and life-tables. Journal of
the royal statistical society. \emph{Series B (Methodological)}
\textbf{34}, 187--220 (1972).}

\bibitem[\citeproctext]{ref-kaplan1958nonparametric}
\CSLLeftMargin{40. }%
\CSLRightInline{Kaplan, E. L. \& Meier, P. Nonparametric estimation from
incomplete observations. \emph{Journal of the American statistical
association} \textbf{53}, 457--481 (1958).}

>>>>

\emph{Manuscript last updated: 2025-07-26 11:39:19.945045}

@@

However, the comparability and transportability of results derived
between non-interventional studies and RCTs depends on multiple factors.
Frequently referenced limitations include lack of baseline randomization
and the imbalance in prognostic factors resulting thereof, missing data,
unmeasured confounding, small study sizes, data
discontinuity,\textsuperscript{3,4} adoption of changes in treatment
guidelines over time in routine clinical care and the inability to
measure and emulate common eligibility criteria, including prognostic
factors, and standardized endpoint assessments in real-world data
(RWD).\textsuperscript{5} While examples of oncology trial emulations
have been published,\textsuperscript{5--7} a systematic and scaled
approach to emulate a diverse set of oncology trials with multiple
heterogeneous databases would enhance increase confidence in the
interpretability of non-interventional studies, evaluate regulatory
considerations, and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{8} evaluated when
non-interventional studies can come to causal conclusions on treatment
effects by benchmarking results against RCTs under the assumption that
well-designed and conducted RCT findings reflect causal treatment
effects. In settings where the RCT designs could be emulated well,
non-interventional studies were able to reach similar
conclusions.\textsuperscript{9} However, this prior work from
RCT-DUPLICATE focused on emulating trials in non-oncology settings using
claims databases.

@@

The focus of ENCORE is to clarify when non-interventional studies can or
cannot yield similar results compared to oncology RCTs. Therefore, the
project emphasizes trials of therapies for common cancers for which
there has been substantial therapeutic development in recent years.
After review and collaboration with clinical and regulatory experts,
four cancer types were identified: non-small cell lung cancer, breast
cancer, colorectal cancer and multiple myeloma. For each cancer type, we
aim to conduct three trial emulations using multiple accessible
databases (i.e., the total equaling 12 trial emulations x \emph{n}
databases which are found fit-for-purpose for each trial).

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: Parallel assignment
\item
  Industry-sponsored
\item
  Trial start: 2011 or later
\item
  Primary purpose: Study treatment efficacy
\item
  Endpoint(s): Overall survival must be one of the endpoints reported
  (either as hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or trials that challenged existing treatment
  policies)
\end{itemize}

The operationalization for each criterion is listed in detail in
Table~\ref{tbl-criteria}. We mainly considered pivotal large late-phase
RCTs that were initiated after 2011 because treatment guidelines for
included cancer indications have undergone significant changes in recent
years. Due to the rapid adoption of newly-approved therapies in routine
care, patients are less likely to receive outdated treatment regimens in
clinical practice. Conversely, we excluded trials with results published
too recently to allow for substantive medical product uptake and
follow-up time accrual in real-world databases used for this project. We
did not define a global cut-off as the requirements for follow-up time
are different for each cancer type and population (e.g., advanced NSCLC
versus early-stage breast cancer) and the decisions were made on a
clinically relevant basis.

@@

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
up candidate trials are listed in \textbf{Supplementary Table 1}. The
majority of shortlisted trials represent indications for advanced
(locally advanced, inoperable, recurrent/progressive disease) or
metastatic cancer patients because a large proportion of drug
development efforts have recently focused on these settings. A key
objective that we aim to explore with the selected trials is to evaluate
how stage (early, late), line of therapy ({[}neo{]}adjuvant, first line,
advanced lines of therapy), therapy protocols (monotherapy, combination
therapy) and clinical characteristics (all-comer versus complex genetic
or immunological signatures) can be emulated using RWD. If more thorough
feasibility assessments after applying all relevant eligibility criteria
(including formal power calculations and assessment of sufficient
balance after propensity score matching or weighting) suggest that the
threat of bias from mis-measurement of key study parameters or residual
confounding remains high, or that the study size is not sufficient, then
runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific
EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA,
Flatiron Health, and Ontada/McKesson. A detailed description and
sampling methodology will be provided with each trial emulation
protocol. For ENCORE, not all databases will be available for each
cancer indication given the specificity of data elements required and
the names of the databases will be blinded and referred to as ENCORE
DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in
randomly assigned order). If more than one database is considered
fit-for-use for a respective trial emulation, the most suitable analytic
model will be employed for each database separately and final treatment
effect estimates will be pooled using a meta-analytic approach using
random effects (primary; weights reflect the inverse of the variance of
each database study plus an additional variance term that quantifies the
assumed heterogeneity between databases) and fixed effects (secondary;
weights reflect the inverse of the variance of each database study)
models.\textsuperscript{20}

@@

The distribution of patient characteristics, by exposure status, will be
examined before and after applying eligibility criteria and contrasted
with the distributions of patient characteristics of the original RCT.
Initial propensity score matching or weighting methods will be applied
to ensure that measured pre-exposure covariates can be
balanced\textsuperscript{28}, exposure cohorts are conditionally
exchangeable at baseline, and resulting sample sizes are still
sufficient after matching or weighting. At this stage, all exploratory
analyses will be conducted blinded towards the outcome to not bias any
study design or analytic choices based on known outcome information.

@@

Causal analyses of non-interventional data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data .\textsuperscript{29} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreement metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time from index treatment initiation to all-cause mortality (rwOS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess whether the overall number
of events, unstratified by exposure, is sufficient such that a
significant difference can be detected based on the original
RCT-reported hazard ratio (HR), the statistical power (1-\(\beta\) with
a 2-sided \(\alpha\) of 0.05) will be estimated using Schoenfeld's
sample-size formula for the proportional-hazards regression
model.\textsuperscript{30}

@@

To formally compare treatment effects between RCTs and their respective
emulations, we will adapt the approach of the RCT-DUPLICATE
project.\textsuperscript{8,31} That is, for the primary endpoint of
interest (hazard ratio {[}HR{]} for OS and corresponding 95\% confidence
intervals), we will derive three qualitative agreement metrics:
statistical agreement, estimate agreement and agreement based on the
standardized mean difference (SMD). Examples are illustrated in
Table~\ref{tbl-metrics}.

@@

where \(\hat{\theta}_{RCT}\) and \(\hat{\theta}_{RWE}\) are the
treatment effect estimates (log hazard ratios or median survival time
differences) and \(\text{Var}(\hat{\theta}_{RCT})\) and
\(\text{Var}(\hat{\theta}_{RWE})\) are the corresponding variances for
RCT and RWE, respectively. The resulting SMDs will be interpreted such
that with an SMD of 1.00, the effect estimate from the RCT and the RWE
emulation are 1 standard deviation apart. For an \(\alpha\)-level of
0.05, the null hypothesis of no difference would be rejected whenever
\(|Z| > 1.96\).

@@

The study design for each trial emulation will be visualized as part of
the protocol using a graphical depiction of the exact measurement
windows of eligibility criteria, washout periods, and covariates
relative to the cohort entry time.\textsuperscript{32}

@@

To establish an analytic cohort, missingness will be assessed across
patients who meet all eligibility criteria, including those with missing
values as a first iteration. These missing data investigations will
empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{33} We will adopt a
principled process on missing data that empirically evaluates different
aspects across partially observed covariates based on three group
diagnostics.\textsuperscript{34,35} These diagnostics cover (1)
comparisons of patients characteristics with and without an observed
level of the partially observed covariate, (2) ability to predict
missingness given observed data, and (3) assessments if outcomes between
patients with a missing value are systematically different. Expert
domain knowledge and assumptions about the underlying missing data
structure through canonical causal diagrams\textsuperscript{36} will
additionally inform decisions regarding the inclusion or exclusion of
individuals with missing values in key eligibility criteria and
potential sensitivity analyses to assess the robustness of these
decisions.\textsuperscript{37}

While MAR is a strong assumption to hold across all considered
covariates, it has been shown that especially in the context of
partially observed covariate data (as opposed to missing exposure and
outcome data), generally only mechanisms in which a covariate causes its
own missingness lead to significant bias (MNAR).\textsuperscript{36}
Hence, methodologies which retain otherwise eligible patients and give
the potential to adjust for a broader set of prognostic factors (e.g.,
multiple imputation\textsuperscript{38} or doubly robust
methods\textsuperscript{39}) may be preferred over complete case
analyses, although such tradeoffs must be carefully evaluated within the
clinical and study contexts.

@@

Due to its frequency of use in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal HR coefficient for
the treatment comparison for OS.\textsuperscript{40} We will also
consider alternative endpoints on an absolute risk scale such as median
survival times, survival probabilities at pre-defined time points during
follow-up\textsuperscript{41} and restricted mean survival times as
secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of relevant prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{42} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{43,44} As appropriate, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{45,46} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{43}

To assess the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is often considered
reasonable at a SMD \textless{} 0.1.\textsuperscript{28} Further, we
will compute the average post-matching or post-weighting
C-statistics.\textsuperscript{47} In addition, we will use a published
prognostic score for OS\textsuperscript{42} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{48}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{41} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given follow-up
time during.\textsuperscript{49,50} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

@@

Building on an established approach developed by the RCT DUPLICATE
project\textsuperscript{8,31}, the ENCORE project aims to emulate 12
oncology trials using four oncology EHR data sources to inform the
potential use of RWD for regulatory purposes in oncology. The project
focuses on four common cancers and will evaluate the agreement of
treatment effect estimates between RCTs and their respective emulations.
Historically, administrative claims databases have been the predominant
data source used to evaluate medical product safety and effectiveness in
the postmarketing setting. With increasing access to EHR data and
availability of granular clinical data elements often necessary to
gather fit-for-use data in oncology, as well as a maturing set of
methodological approaches for causal inference using such data, the
ENCORE project is one of the first of its kind to evaluate contexts
where RWD may be reliably used to draw similar conclusions compared to
RCTs in the field of oncology.

The RCT DUPLICATE initiative\textsuperscript{8,31} identified multiple
emulation challenges that may be similar in ENCORE. One particular
aspect that we may not always be able to emulate is the exact
distribution of patient characteristics of the trial population. This
can be due to the lack of data granularity and comprehensiveness to
emulate relevant eligibility criteria or the fact that large pivotal
trials in oncology are typically conducted in multiple countries
worldwide (all considered databases reflect the US only). This may
impact the results of this trial emulation given that the distribution
of genetic alterations and biomarkers, prognosis, and factors that drive
heterogeneous treatment outcomes of certain cancers can differ between
countries. For example, certain cancers (e.g., GI cancers) or certain
genetic mutations (e.g., EGFR mutations in lung cancer) are much more
prevalent in certain geographic regions and populations compared to the
US. Additionally, treatment landscapes such as the preference for
specific regimens or compounds, their use in earlier or later lines of
therapies, and supportive care practices can differ between geographic
regions, too.

@@

\bibitem[\citeproctext]{ref-austin2009balance}
\CSLLeftMargin{28. }%
\CSLRightInline{Austin, P. C. Balance diagnostics for comparing the
distribution of baseline covariates between treatment groups in
propensity-score matched samples. \emph{Statistics in medicine}
\textbf{28}, 3083--3107 (2009).}

\bibitem[\citeproctext]{ref-hernan2022causal}
\CSLLeftMargin{29. }%
\CSLRightInline{Hernán, M. A. Causal analyses of existing databases: No
power calculations required. \emph{Journal of clinical epidemiology}
\textbf{144}, 203--205 (2022).}

\bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
\CSLLeftMargin{30. }%
\CSLRightInline{Schoenfeld, D. The asymptotic properties of
nonparametric tests for comparing survival distributions.
\emph{Biometrika} \textbf{68}, 316--319 (1981).}

\bibitem[\citeproctext]{ref-franklin2020nonrandomized}
\CSLLeftMargin{31. }%
\CSLRightInline{Franklin, J. M. \emph{et al.} Nonrandomized real-world
evidence to support regulatory decision making: Process for a randomized
trial replication project. \emph{Clinical Pharmacology \& Therapeutics}
\textbf{107}, 817--826 (2020).}

\bibitem[\citeproctext]{ref-schneeweiss2019graphical}
\CSLLeftMargin{32. }%
\CSLRightInline{Schneeweiss, S. \emph{et al.} Graphical depiction of
longitudinal study designs in health care databases. \emph{Annals of
internal medicine} \textbf{170}, 398--406 (2019).}

\bibitem[\citeproctext]{ref-rubin1976inference}
\CSLLeftMargin{33. }%
\CSLRightInline{Rubin, D. B. Inference and missing data.
\emph{Biometrika} \textbf{63}, 581--592 (1976).}

\bibitem[\citeproctext]{ref-weberpals2024smdi}
\CSLLeftMargin{34. }%
\CSLRightInline{Weberpals, J. \emph{et al.} Smdi: An r package to
perform structural missing data investigations on partially observed
confounders in real-world evidence studies. \emph{JAMIA open}
\textbf{7}, ooae008 (2024).}

\bibitem[\citeproctext]{ref-weberpals2024}
\CSLLeftMargin{35. }%
\CSLRightInline{Weberpals, J. \emph{et al.}
\href{https://doi.org/10.2147/CLEP.S436131}{A principled approach to
characterize and analyze partially observed confounder data from
electronic health records}. \emph{Clinical Epidemiology} \textbf{16},
329--343 (2024).}

\bibitem[\citeproctext]{ref-moreno2018canonical}
\CSLLeftMargin{36. }%
\CSLRightInline{Moreno-Betancur, M. \emph{et al.} Canonical causal
diagrams to guide the treatment of missing data in epidemiologic
studies. \emph{American journal of epidemiology} \textbf{187},
2705--2715 (2018).}

\bibitem[\citeproctext]{ref-tompsett2018use}
\CSLLeftMargin{37. }%
\CSLRightInline{Tompsett, D. M., Leacy, F., Moreno-Betancur, M., Heron,
J. \& White, I. R. On the use of the not-at-random fully conditional
specification (NARFCS) procedure in practice. \emph{Statistics in
medicine} \textbf{37}, 2338--2353 (2018).}

\bibitem[\citeproctext]{ref-weberpals2024hdmi}
\CSLLeftMargin{38. }%
\CSLRightInline{Weberpals, J. \emph{et al.} High-dimensional multiple
imputation (HDMI) for partially observed confounders including natural
language processing-derived auxiliary covariates. \emph{American Journal
of Epidemiology} (2025).at
\textless{}\url{https://arxiv.org/abs/2405.10925}\textgreater{}}

\bibitem[\citeproctext]{ref-Shaw2024}
\CSLLeftMargin{39. }%
\CSLRightInline{Shaw, C. K., P. \& Williamson, B. D. Assessing treatment
effects in observational data with missing confounders: A comparative
study of practical doubly-robust and traditional missing data methods.
\emph{GitHub repository} (2024).at
\textless{}\url{https://github.com/PamelaShaw/Missing-Confounders-Methods}\textgreater{}}

\bibitem[\citeproctext]{ref-cox1972regression}
\CSLLeftMargin{40. }%
\CSLRightInline{Cox, D. R. Regression models and life-tables. Journal of
the royal statistical society. \emph{Series B (Methodological)}
\textbf{34}, 187--220 (1972).}

\bibitem[\citeproctext]{ref-kaplan1958nonparametric}
\CSLLeftMargin{41. }%
\CSLRightInline{Kaplan, E. L. \& Meier, P. Nonparametric estimation from
incomplete observations. \emph{Journal of the American statistical
association} \textbf{53}, 457--481 (1958).}
